---
title: "Final Report"
author: "Helen Nguyen, Tejasvi Konakanchi, Sruti Banerjee"
output: html_document
---

## Abstract
This project investigates the relationship between socioeconomic factors and police violence in the United States. Using datasets related to “Police Killings”, “Median Household Income”, “Population Statistics”, “Poverty Rates” we analyzed the patterns in police brutality in 2015. Using visualization methods such as scatter plots, bar plots and heat maps we were able to answer and analyze the issue. Our observations from data visualization:

* As median household income (per state) increases, there is a decrease in police violence.
* As the poverty rate (across states) increases, there is an increase in violent police incidents.
* Larger population regions tend to have higher average poverty rates.
* The state with highest police violence is ‘New Mexico’. 

Data pre-processing techniques including standardizing, cleaning and merging have been performed to ensure accuracy. Correlation analysis and statistical analysis have been applied to confirm the relationship between these factors.

## Introduction
Police brutality remains a pressing issue in the United States, characterized by the excessive, unjustified, or unlawful use of force by law enforcement officers. The issue disproportionately affects marginalized communities and raises concerns about systemic bias in law enforcement (The Washington Post, 2015). Understanding the socioeconomic factors linked to police violence is important for addressing these disparities and demanding policy reforms. 

This report seeks to examine several key questions related to police brutality and economic factors. First, what is the relationship between median household income and police killings across the U.S.? This question helps determine whether economic status influences the likelihood of fatal police encounters. Second, do states with higher poverty rates experience higher rates of police killings? Understanding this relationship can shed light on the income disparities in police violence. Third, how is the poverty rate related to regional population size–do larger regions tend to have higher or lower poverty rates? Answering this can help policy makers allocate resources effectively for affected regions. Finally, which states have significantly high rates of police brutality? By identifying these states can help urge local governments to take necessary action to improve public safety. By addressing these questions, we aim to contribute to ongoing discussions on law enforcement accountability and socioeconomic inequality. The findings may serve as a valuable resource for policymakers, law enforcement agencies, and researchers working toward equitable and just policing practices in the United States.

## Data Source
We utilized four datasets to analyze the relationship between socioeconomic factors and police killings in the United States. The [PoliceKillingsUS.csv](https://www.kaggle.com/datasets/kwullum/fatal-police-shootings-in-the-us?select=PoliceKillingsUS.csv), [MedianHouseholdIncome2015.csv](https://www.kaggle.com/datasets/kwullum/fatal-police-shootings-in-the-us?select=MedianHouseholdIncome2015.csv), and [PercentagePeopleBelowPovertyLevel.csv](https://www.kaggle.com/datasets/kwullum/fatal-police-shootings-in-the-us?select=PercentagePeopleBelowPovertyLevel.csv) dataset were sourced on Kaggle and posted by the user Karolina Wullum and last updated 7 years ago.The census data in these datasets were gathered from factfinder.census.gov and the police killings data set was compiled by The Washington Post. Additionally, we used [nst-est2019-01.xlsx](https://www.census.gov/data/tables/time-series/demo/popest/2010s-national-total.html#par_textimage), which was collected from the United States census bureau under “State Population totals 2010-2019.” We only extracted out the population data for 2015. 

Each dataset provides key insights into our research questions. The [PoliceKillingsUS.csv](https://www.kaggle.com/datasets/kwullum/fatal-police-shootings-in-the-us?select=PoliceKillingsUS.csv) dataset documents individual cases of fatal police shootings in each U.S. city in  2015. [MedianHouseholdIncome2015.csv](https://www.kaggle.com/datasets/kwullum/fatal-police-shootings-in-the-us?select=MedianHouseholdIncome2015.csv) represents the median household income by each state while [PercentagePeopleBelowPovertyLevel.csv](https://www.kaggle.com/datasets/kwullum/fatal-police-shootings-in-the-us?select=PercentagePeopleBelowPovertyLevel.csv) captures the percentage of people who are under the poverty line in each city in the states. [Nst-est2019-01.xlsx](https://www.census.gov/data/tables/time-series/demo/popest/2010s-national-total.html#par_textimage) contains the estimated state population over time. We selected these datasets because they align with our research objectives, helping us explore how economic status, poverty levels, and regional demographics may influence police violence. By analyzing these factors, we aim to see the potential socioeconomic patterns in police killings and identify communities most affected by it.

## Data

**MedianHouseholdIncome2015.csv:**

* Number of rows: 29322, Number of columns: 3
* Each row represents a specific city or town (region) in the United States of America, along with the corresponding geographic area (state).
* Relevant variables and their coding: 
    * _Geographic Area_: A two letter state abbreviation.
    * _City_: The name of the city or town.
    * _Median Income_: The median household income for that city or town, which appears to be stored as a string but represents numeric values.
* There are 51 missing values in the _Median Income_ variable, the missing value is coded as “(X)”.


**PoliceKillings.csv:**

* Number of rows: 2535, Number of columns: 14
* Each row represents an individual harmed by police in the US.
* Relevant variables and their coding: 
    * _id_: Unique identifier for each case. 
    * _name_: Name of the individual killed. 
    * _date_: Date of the incident. 
    * _manner_of_death_: How the person was killed (e.g., "shot", "shot and Tasered"). 
    * _armed_: Indicates whether the individual was armed and with what (e.g., "gun", "knife", "unarmed"). 
    * _age_: Age of the individual. 
    * _gender_: Gender of the individual (M for male, F for female). 
    * _race_: Race of the individual (e.g., "W" for White, "B" for Black, "H" for Hispanic). 
    * _city_: City where the incident occurred. 
    * _state_: State where the incident occurred. 
    * _signs_of_mental_illness_: Whether there were signs of mental illness (True or False). 
    * _threat_level_: Perceived threat level (attack, other, undetermined). 
    * _flee_: Whether the individual was fleeing (Foot, Car, Not fleeing, etc.). 
    * _body_camera_: Whether a body camera was worn (True or False).
* Missing Values: 
    * _armed_: 9 missing values 
    * _age_: 77 missing values 
    * _race_: 195 missing values 
    * _flee_: 65 missing values

**PercentagePeopleBelowPovertyLevel.csv:**

* Number of rows: 29329, Number of columns: 3
* Each row represents a specific city or town in the United States, along with its corresponding state and the percentage of people living below the poverty level in that area.
* Relevant variables and their coding: 
    * _Geographic Area_: The two-letter state abbreviation (e.g., AL for Alabama). 
    * _City_: The name of the city or town. 
    * _poverty_rate_: The percentage of people living below the poverty level in that city.
* Missing Values: 
    * _Poverty_rate_: 201 entries, the missing value is coded as non numeric variables such as “-”

**Nst-est2019-01.xlsx:**

* Number of rows: 6 rows, Number of columns
* Each row represents each state’s estimated population for each year from 2010-2019
* Relevant variables and their coding:
    * _state_: Name of the state associated with the estimated population
    * _population_: the estimated population for each state
* Missing Values:
    * _population_: 96 missing values
    
## Methods
We renamed longer column names like “Geographic Area”, “Median Income”, etc. to shorter names like “state”, “popEst2015”, and “median_income” to make it easier to code. We filtered out all the rows with N/A values from all datasets. We used “state” as the merge key to combine all datasets.

For our poverty dataset, we calculated the average poverty rate across all cities in a state and used that average as an indicator of the poverty rate for a state. For our population dataset, we got rid of the data concerning US regions and only kept the data concerning each individual state. For our police killings dataset, we calculated the total police killings across a state and used that as an indicator of police brutality in a state.

We decided to only look at data from 2015, as all our datasets had observations for that year. The relevant variables for our questions were state and the poverty rate, median income, police killings, and police killings per 100k residents for each state.

Our biggest challenge was the vast amount of data we had. It was difficult figuring out how to best display data for all 50 states, especially for research questions 3 and 4. We finally decided to split the states into 4 different regions for question 3, and to use a heat map for question 4.

```{r echo=FALSE, message=FALSE, results='hide'}
library(tidyverse)
library(readxl)

## Loading data
MHI <- read_delim("MedianHouseholdIncome2015.csv", show_col_types = FALSE) # Median Household Income
PK <- read_delim("PoliceKillingsUS.csv", show_col_types = FALSE) # Police Killings
Pop <- read_excel("nst-est2019-01.xlsx") # Population by state
PR <- read_delim("PercentagePeopleBelowPovertyLevel.csv", show_col_types = FALSE) #Poverty rate
```
```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## PR cleaning & reshaping
colnames(PR)

names(PR)[names(PR) == "Geographic Area"] <- "state"

PR_clean <- PR %>%
  select(state, poverty_rate) %>%
  filter(!is.na(state) & !is.na(poverty_rate)) %>%
  group_by(state) %>%
  mutate(poverty_rate = as.numeric(poverty_rate)) %>%
  summarize(avg_poverty_rate = mean(poverty_rate, na.rm = TRUE))
head(PR_clean)

min(PR_clean$avg_poverty_rate)
max(PR_clean$avg_poverty_rate)
mean(PR_clean$avg_poverty_rate)

```


```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## renaming the two columns for Pop data set
names(Pop)[names(Pop) == "table with row headers in column A and column headers in rows 3 through 4. (leading dots indicate sub-parts)"] <- "state"
names(Pop)[names(Pop) == "...9"] <- "popEst2015"
Pop$state <- str_remove(Pop$state, "^\\.") # Gets rid of the "." in the data for States


# Gets rid of the first few rows & only shows states abbreviations and population during 2015
Pop <- Pop %>%
  select(state, popEst2015) %>%
  slice(-(1:8)) %>%
  mutate(state = state.abb[match(state, state.name)], 
                            state = replace(state, 9, "DC"))

## Pop has missing values and have been cleaned
Pop_clean <- Pop %>%
    filter(!is.na(state) & !is.na(popEst2015))
nrow(Pop_clean)
colnames(Pop)
head(Pop_clean)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## MHI has missing income values which have been cleaned.
MHI_clean <- MHI %>%
  filter(!is.na(`Median Income`) & `Median Income` != "(X)" & `Median Income` != "-")

## changing column name for MHI
names(MHI_clean)[names(MHI_clean) == "Geographic Area"] <- "state"
names(MHI_clean)[names(MHI_clean) == "Median Income"] <- "median_income"

MHI_clean <- MHI_clean %>%
  select(state, median_income) %>%
  group_by(state) %>%
  mutate(median_income = as.numeric(median_income)) %>%
  summarize(median_income = mean(median_income, na.rm = TRUE))

head(MHI_clean)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## PK has been cleaned from missing values and has the sum of killings per state
PK_clean <- PK %>%
  select(state) %>%
  filter(!is.na(state)) %>%
  group_by(state) %>%
  summarize(PoliceKillings = n())

head(PK_clean)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## Merging
groupDataSet <- PK_clean %>%
  full_join(MHI_clean, by = "state") %>%
  full_join(Pop_clean, by = "state") %>%
  full_join(PR_clean, by = "state")

head(groupDataSet)
```

## Results
### RQ 01: What is the relationship between median household income and police killings across the States in US?

From our data visualization, it shows that there is a negative correlation. So, we can say that as median household income increases, police killings per 100K residents decreases. Our scatter plot shows a downward trend reinforcing the negative correlation. The blue trend line has a negative slope, meaning high-income states generally experience fewer violent police incidents per capita. This suggests that socioeconomic factors may influence the rate of police brutality. However, when calculating Pearson's correlation, p = 0.007 indicates significant statistical relevance, meaning that there is no relationship between median household income and police killings across the U.S..

**Note:** A log transformation was applied to handle skewness in the data. The log reduced the influence of extreme values and made the relationship more interpretable.

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## RQ01
groupDataSetRQ1 <- groupDataSet %>%
  mutate(popEst2015 = as.numeric(popEst2015),
         killings_per_100k = (PoliceKillings / popEst2015) * 100000)%>%
  mutate(log_killings_per_100k = log(killings_per_100k + 1))

# Plotting the merged data sets to answer RQ01
ggplot(groupDataSetRQ1, aes(x = median_income, y = log_killings_per_100k)) +
  geom_point(color = "yellow2", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "lightblue3") + 
  labs(
    title = "Log(Violence per 100k Residents) vs. Median Income on State Level (2015)",
    x = "Average Median Household Income (USD)",
    y = "Log(Violence per 100k Residents)"
  ) +
  theme_minimal()
```


### RQ 02: Do states with higher poverty rates experience higher rates of police killings?

Our scatter plot shows an upward trend reinforcing the positive correlation. The statistically significant positive correlation suggests a meaningful relationship between these variables. The gold trend line has a positive slope, meaning as the average poverty increases states generally experience higher violent police incidents per capita. However, when calculating the Pearson correlation for this plot, we found p = 0.016, indicating that there is no relationship between high poverty rates and high rates of police killings.

**Note:** A log transformation was applied to handle skewness in the data. The log reduced the influence of extreme values and made the relationship more interpretable.

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## RQ2
groupDataSetRQ2 <- groupDataSet %>%
  mutate(popEst2015 = as.numeric(popEst2015),
         killings_per_100k = (PoliceKillings / popEst2015) * 100000)%>%
  mutate(log_killings_per_100k = log(killings_per_100k + 1)) %>%
  
  select(state, PoliceKillings, 
         avg_poverty_rate, 
         killings_per_100k, 
         log_killings_per_100k)

## Plotting the poverty and police killings
ggplot(groupDataSetRQ2, aes(x = avg_poverty_rate, y = killings_per_100k)) +
  geom_point(alpha = 0.4, color = "purple") +
  geom_smooth(method = "lm", se = TRUE, color = "gold2") + 
  labs(
    title = "Log(Violence per 100k Residents) vs. Average poverty rate (2015)",
    x = "Average poverty rate",
    y = "Log(Violence per 100k Residents)"
  ) +
  theme_minimal()


```
```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
#testing co-relation
cor.test(groupDataSetRQ2$avg_poverty_rate, groupDataSetRQ2$log_killings_per_100k, method = "pearson")

```

### RQ 03: How is the poverty rate related to the region population size? Do larger regions tend to have higher or lower poverty rates?

On the bar graph, we see that the southern region holds the largest population while also having the highest poverty rate. Indicating that there is a relationship between population size in a region and poverty rates. However, the second largest region is the northeast region and they have the lowest rate of poverty. This shows that although there may be a relationship between regional population size and poverty rates, there isn’t a strong correlation to prove it.

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## RQ3
groupDataSetRQ3 <- groupDataSet %>%
  select(state, popEst2015, avg_poverty_rate) %>%
  mutate(
    region = case_when(
      state %in% c("ME", "NH", "VT", "MA", "RI", "CT", "NY", "NJ", "PA") ~ "NE",  # Northeast
      state %in% c("OH", "IN", "IL", "MI", "WI", "MO", "IA", "MN", "ND", "SD", "NE", "KS") ~ "MW",  # Midwest
      state %in% c("DE", "MD", "DC", "VA", "NC", "SC", "GA", "FL", "KY", "TN", "AL", "MS", "AR", "LA", "OK", "TX", "WV") ~ "S",  # South
      state %in% c("MT", "WY", "CO", "ID", "UT", "NV", "AZ", "NM", "AK", "WA", "OR", "CA", "HI") ~ "W"  # West
    )
  ) %>%
  group_by(region) %>%
  summarise(
    avg_poverty_rate_region = mean(avg_poverty_rate, na.rm = TRUE),  # Average poverty rate per region
    avg_population_region = mean(popEst2015, na.rm = TRUE)  # Average population per region
  ) %>%
  mutate(avg_population_region = as.numeric(avg_population_region),
         avg_poverty_rate_region = as.numeric(avg_poverty_rate_region))

groupDataSetRQ3_long <- groupDataSetRQ3 %>%
  pivot_longer(cols = c(avg_poverty_rate_region, avg_population_region),
               names_to = "metric",
               values_to = "value")

ggplot(groupDataSetRQ3_long, aes(x = region, y = value, fill = region)) +
  geom_col() +
  facet_wrap(~metric, scales = "free_y") +
  labs(
    title = "Average Poverty Rate and Population Size per Region",
    x = "Region",
    y = "Value"
  ) +
  theme_minimal()

```

### RQ 04: Is there a specific area that has significantly high police brutality in the US?

From our data visualization, we can see that New Mexico has a significantly high police brutality rate per 100k residences with Alaska and Oklahoma tailing behind. This indicates that these states are the hotspots and have a significantly higher police brutality rate in the U.S.

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
## RQ4
library(usmap)
library(viridis)
groupDataSetRQ4 <- groupDataSet %>%
  mutate(popEst2015 = as.numeric(popEst2015),
         killings_per_100k = (PoliceKillings / popEst2015) * 100000)

plot_usmap(data = groupDataSetRQ4, values = "killings_per_100k", regions = "states") +
  scale_fill_gradientn(colors = c("#F9FBF2", "#FFEDE1", "#D7F9FF", "#AFCBFF", "#0E1C36"), name = "Police Killings") +
  theme_minimal() +
  labs(title = "Police Killings per 100k Residents by State in the U.S.",
       subtitle = "Darker states indicate higher police killings")
```

## Discussion

Our findings provide strong evidence that socioeconomic disparities and police violence are related. We observed a negative correlation between median household income and police killings; this suggests that wealthier states experience lower levels of police violence per capita. We saw a positive correlation between the percentage of people below poverty level and police killings, this proves that as the percentage of people below poverty rate in a state increases, the police violence experienced by the state also increases. We also observed that regions with higher populations typically have a higher percentage of people below poverty level. We also observed that New Mexico is the state that experienced the highest police violence. These observations between poverty and police killings reinforces the notion that regions facing economic hardships are more likely to be involved in protests, thefts, or burglary which would result in a risk of experiencing fatal police encounters.

Larger populations tend to have higher average poverty rates. These findings align with broader socioeconomic trends where urban centers with higher populations often have significant income disparities, contributing to systemic issues such as crime, policing, and community-police interactions. The identification of New Mexico as having the highest police violence per capita raises concerns about localized law enforcement policies and disproportionate impact on residents.

**Limitations of the Data:**

* Data limitations: Our dataset only includes police killings record in 2015, although 2015 has been a pivotal moment in the recording of police killings, it limits the generalizability of our findings to different time periods.
* The dataset does not provide detailed individual socioeconomic backgrounds, making it difficult to establish direct causal links between poverty and police violence on an individual scale.
* The data relies on sources such as the Washington Post, which may not capture all instances of police killings, especially those occurring in underreported on unmonitored areas.
* While our analysis focuses on state-level trends, variations at the city or county level may present different patterns that were not properly captured in this study.

**Directions for Future Research:**

* To build up on these findings future research should:
    * Analyze police killings over multiple years to help determine trends or specific policy changes that caused social movements and a result spike in violence experienced by the residents.
    * Including variables such as education levels, crime rates, and racial demographics for future analyses could provide a more comprehensive understanding of the factors influencing police violence.
    * Investigating city level data would help identify urban areas where policy reforms may be most needed.
    * Exploring how different policing policies such as body cameras, de-escalation training affect police violence rates, could provide actionable insights for policy makers.

By addressing these limitations and extending our research scope, we can contribute to a more nuanced understanding of the socioeconomic factors and police violence and make a meaningful policy change to promote safer and more equitable communities.


## Summary

Our findings are further proof of how marginalized communities are disproportionately affected by police violence. We are doing a disservice to the most vulnerable people of our population. We hope that these findings urge policymakers, researchers, and activists to brainstorm ways to improve policing such that marginalized groups are no longer harmed by the very system that is meant to protect them.

